rOpenSci tools for textmining open source science literature

### 2017-09-13 manuscript.Rmd
updates to manuscript, added new sections

Explosion of digital context online. But we need tools to facilitate textmining (i.e., rOpenSci)

Why text mine it.

How to text mine it.

Here, we introduce and give examples for textmining in R using rOpenSci packages, and related packages.

# Text mining sources

[Describe open source data for text mining that we link to in R, and those we don't link to in R.]

There is increasing open source scientific literature content available online. However, only a small proportion of scientific journals provide access to their full content; whereas, many publishers provide open access to their metadata only (Table 1).

#### Table 1. Sources of scientific literature, their content type provided via web services, whether rOpenSci has an R packages for the service, and where to find the API documentation.

Data Provider                      | Content Type            | rOpenSci? | API Documentation
---------------------------------- | ---------------------   | --------- | -----------------
Public Library of Science (PLoS)   | Full text/altmetrics    | rplos     | http://api.plos.org/
Springer                           | Full text on OA content | rspringer | http://dev.springer.com/
Pensoft                            | Full text/altmetrics    | rpensoft  | http://bit.ly/KYP0Zi
Nature Publishing Group            | Metadata only           | No        | http://developers.nature.com/
Mendeley                           | Metadata only           | rmendeley | http://dev.mendeley.com/
DataCite                           | Metadata only           | rdatacite | http://oai.datacite.org/
Biodiversity Heritage Library      | Full content            | rbhl      | http://bit.ly/KYQ1Rd
Scopus (Elsevier)                  | Metadata only ????      | No        | http://bit.ly/J9S616

#### Table 2. Tools for text mining science literature.

Tool           |  Platform            | Services              | URL
-------------- | -------------------- | --------------------- | ---------------------
PMC Miner      | Python               | PubMed                | http://bit.ly/L29ekY
rOpenSci       | R                    | See Table 1           | http://ropensci.org/
PMC Miner      | Python               | PubMed                | http://bit.ly/L29ekY
PMC Miner      | Python               | PubMed                | http://bit.ly/L29ekY
PMC Miner      | Python               | PubMed                | http://bit.ly/L29ekY
PMC Miner      | Python               | PubMed                | http://bit.ly/L29ekY
PMC Miner      | Python               | PubMed                | http://bit.ly/L29ekY
PMC Miner      | Python               | PubMed                | http://bit.ly/L29ekY
PMC Miner      | Python               | PubMed                | http://bit.ly/L29ekY
PMC Miner      | Python               | PubMed                | http://bit.ly/L29ekY
PMC Miner      | Python               | PubMed                | http://bit.ly/L29ekY
PMC Miner      | Python               | PubMed                | http://bit.ly/L29ekY

@@

@@

>>>>

There's likely more than 100 million articles published
(source: Crossref API), representing an enormous amount
of knowledge.  In addition to simply reading these articles,
they contain a vast trove of information of interest to
researchers.

For example, many researchers are interested in statistical
outcomes of articles: questions about P-values, about effect
sizes, and more. With regard to effect sizes, these are
of particular interest, as they are often combined in
meta-analyses to draw broad conclusions about a particular
question.

Text-mining is the broad term associated with pulling information
out of articles. Given the importance of text-mining, good
text-mining tools are needed to make it easier for researchers
to do.

Here, we do an overview of text-mining tools in the R programming
language. We do not cover analysis tools per se, but rather those
tools for searching for, acquiring, and "mashing up" text.

# Digital articles: technical aspects

Of digital articles some of which are available digitally,
and some of which are not. Those that are digital can be split
into two groups: easily machine readable and non-machine readable.

The machine readable articles are those in XML, JSON, or plain text
format. The former two, XML and JSON, are ideal of the machine
readable types because they are structured data, whereas
plain text has no structure - it's simply a long set of characters
with line breaks and spaces in between.

Of the non-machine readable kind, there's PDFs. These can be broken out
into two groups: text based PDFs and scanned PDFs. The former are
converted from digital versions of various kinds (MS Word, OpenOffice,
markdown, etc.), while the latter are PDFs created by scanning
in print articles for which there is no digital version.

# Digital articles: the access landscape

Acces to full-text is the holy grail in text-mining. Some use cases can
get by with article metadata (authors, title, etc.), some with abstracts,
but many use cases need full-text.

The landscape of access to full-text is a extremely hetergeous, with the
majority of variation along the publisher axis. The major hurdle are
paywalls. The majority of articles are published by the big three
publishers - Wiley, Springer, Elsever - and the majority of their articles
are behind paywalls.

A promising sign is that there's an increasing number of open access
publishers. xxxx.

# The discovery problem (maybe remove section)

xxx

# Data sources

There is increasing open source scientific literature content available
online. However, only a small proportion of scientific journals provide
access to their full content; whereas, most publishers provide open
access to their metadata only (most often through Crossref; Table 1).

Table 1. Sources of scientific literature, their content type provided via web services,
whether rOpenSci has an R packages for the service, and where to find the API documentation.

Data Provider                      | Content Type            | rOpenSci Pkg?            | API Documentation
---------------------------------- | ---------------------   | ------------------------ | -----------------
Crossref                           | Metadata only           | rcrossref                | [^1]
DataCite                           | Metadata only           | rdatacite                | [^2]
Biodiversity Heritage Library      | Full content/Metadata   | rbhl                     | [^3]
Public Library of Science (PLoS)   | Full text/altmetrics    | rplos                    | [^4]
Scopus (Elsevier)                  | Full content/Metadata   | fulltext                 | [^5]
arXiv                              | Full content/Metadata   | aRxiv                    | [^6]
Biomed Central (via Springer)      | Full content/Metadata   | fulltext                 | [^7]
bioRxiv                            | Full content/Metadata   | fulltext                 | [^8]
PMC/Pubmed (via Entrez)            | Full content/Metadata   | rentrez                  | [^9]
Microsoft Academic Search          | Metadata                | fulltext/microdemic      | [^10]

[^1]: http://api.crossref.org
[^2]: https://support.datacite.org/docs/api
[^3]: http://bit.ly/KYQ1Rd
[^4]: http://api.plos.org/solr
[^5]: http://bit.ly/J9S616
[^6]: https://arxiv.org/help/api/index
[^7]: https://dev.springer.com/
[^8]: http://www.biorxiv.org/
[^9]: https://www.ncbi.nlm.nih.gov/books/NBK25500
[^10]: https://azure.microsoft.com/en-us/services/cognitive-services

The following is a synopsis of the major data sources and associated R tools.

## Crossref

Crossref is a non-profit that creates (or "mints") Digital Object Identifiers (DOIs).
In addition, they maintain metadata associated with each DOI. The metadata ranges
from simple (including author, title, dates, DOI, type, publisher) to including
number of citations to the article, as well as references in the article, and even
abstracts.

Crossref does have a text-mining opt-in program for publishers. The result of this is that some
publishers deposit URLs for full text content of their articles. The majority of these links
are pay-walled, while some are open access. Using any of the various tools for working with
Crossref data, you can filter your search to get only articles with full text links, and further
to get only articles with full text links that are open access.

The main interface for Crossref in R is [rcrossref][]. Parallel interfaces are available in
Ruby ([serrano][]) and Python ([habanero][]).

## Pubmed

Pubmed is a corpus/website of NIH funded research ...

@@

xxxx

@@

[rcrossref]: https://github.com/ropensci/rcrossref
[serrano]: https://github.com/sckott/serrano
[habanero]: https://github.com/sckott/habanero

### 2017-09-13 manuscript.tex
updates to manuscript, added new sections

Explosion of digital context online. But we need tools to facilitate
textmining (i.e., rOpenSci)

Why text mine it.

How to text mine it.

Here, we introduce and give examples for textmining in R using rOpenSci
packages, and related packages.

\section{Text mining sources}\label{text-mining-sources}

{[}Describe open source data for text mining that we link to in R, and
those we don't link to in R.{]}

There is increasing open source scientific literature content available
online. However, only a small proportion of scientific journals provide
access to their full content; whereas, many publishers provide open
access to their metadata only (Table 1).

\paragraph{Table 1. Sources of scientific literature, their content type
provided via web services, whether rOpenSci has an R packages for the
service, and where to find the API
documentation.}\label{table-1.-sources-of-scientific-literature-their-content-type-provided-via-web-services-whether-ropensci-has-an-r-packages-for-the-service-and-where-to-find-the-api-documentation.}

\begin{longtable}[]{@{}llll@{}}
\toprule
\begin{minipage}[b]{0.36\columnwidth}\raggedright\strut
Data Provider\strut
\end{minipage} & \begin{minipage}[b]{0.23\columnwidth}\raggedright\strut
Content Type\strut
\end{minipage} & \begin{minipage}[b]{0.10\columnwidth}\raggedright\strut
rOpenSci?\strut
\end{minipage} & \begin{minipage}[b]{0.19\columnwidth}\raggedright\strut
API Documentation\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.36\columnwidth}\raggedright\strut
Public Library of Science (PLoS)\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright\strut
Full text/altmetrics\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
rplos\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
\url{http://api.plos.org/}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.36\columnwidth}\raggedright\strut
Springer\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright\strut
Full text on OA content\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
rspringer\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
\url{http://dev.springer.com/}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.36\columnwidth}\raggedright\strut
Pensoft\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright\strut
Full text/altmetrics\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
rpensoft\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
\url{http://bit.ly/KYP0Zi}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.36\columnwidth}\raggedright\strut
Nature Publishing Group\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright\strut
Metadata only\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
No\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
\url{http://developers.nature.com/}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.36\columnwidth}\raggedright\strut
Mendeley\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright\strut
Metadata only\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
rmendeley\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
\url{http://dev.mendeley.com/}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.36\columnwidth}\raggedright\strut
DataCite\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright\strut
Metadata only\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
rdatacite\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
\url{http://oai.datacite.org/}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.36\columnwidth}\raggedright\strut
Biodiversity Heritage Library\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright\strut
Full content\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
rbhl\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
\url{http://bit.ly/KYQ1Rd}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.36\columnwidth}\raggedright\strut
Scopus (Elsevier)\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright\strut
Metadata only ????\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
No\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
\url{http://bit.ly/J9S616}\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

\paragraph{Table 2. Tools for text mining science
literature.}\label{table-2.-tools-for-text-mining-science-literature.}

\begin{longtable}[]{@{}llll@{}}
\toprule
Tool & Platform & Services & URL\tabularnewline
\midrule
\endhead
PMC Miner & Python & PubMed & \url{http://bit.ly/L29ekY}\tabularnewline
rOpenSci & R & See Table 1 & \url{http://ropensci.org/}\tabularnewline
PMC Miner & Python & PubMed & \url{http://bit.ly/L29ekY}\tabularnewline
PMC Miner & Python & PubMed & \url{http://bit.ly/L29ekY}\tabularnewline
PMC Miner & Python & PubMed & \url{http://bit.ly/L29ekY}\tabularnewline
PMC Miner & Python & PubMed & \url{http://bit.ly/L29ekY}\tabularnewline
PMC Miner & Python & PubMed & \url{http://bit.ly/L29ekY}\tabularnewline
PMC Miner & Python & PubMed & \url{http://bit.ly/L29ekY}\tabularnewline
PMC Miner & Python & PubMed & \url{http://bit.ly/L29ekY}\tabularnewline
PMC Miner & Python & PubMed & \url{http://bit.ly/L29ekY}\tabularnewline
PMC Miner & Python & PubMed & \url{http://bit.ly/L29ekY}\tabularnewline
PMC Miner & Python & PubMed & \url{http://bit.ly/L29ekY}\tabularnewline
\bottomrule
\end{longtable}

>>>>

There's likely more than 100 million articles published (source:
Crossref API), representing an enormous amount of knowledge. In addition
to simply reading these articles, they contain a vast trove of
information of interest to researchers.

For example, many researchers are interested in statistical outcomes of
articles: questions about P-values, about effect sizes, and more. With
regard to effect sizes, these are of particular interest, as they are
often combined in meta-analyses to draw broad conclusions about a
particular question.

Text-mining is the broad term associated with pulling information out of
articles. Given the importance of text-mining, good text-mining tools
are needed to make it easier for researchers to do.

Here, we do an overview of text-mining tools in the R programming
language. We do not cover analysis tools per se, but rather those tools
for searching for, acquiring, and ``mashing up'' text.

\section{Digital articles: technical
aspects}\label{digital-articles-technical-aspects}

Of digital articles some of which are available digitally, and some of
which are not. Those that are digital can be split into two groups:
easily machine readable and non-machine readable.

The machine readable articles are those in XML, JSON, or plain text
format. The former two, XML and JSON, are ideal of the machine readable
types because they are structured data, whereas plain text has no
structure - it's simply a long set of characters with line breaks and
spaces in between.

Of the non-machine readable kind, there's PDFs. These can be broken out
into two groups: text based PDFs and scanned PDFs. The former are
converted from digital versions of various kinds (MS Word, OpenOffice,
markdown, etc.), while the latter are PDFs created by scanning in print
articles for which there is no digital version.

\section{Digital articles: the access
landscape}\label{digital-articles-the-access-landscape}

Acces to full-text is the holy grail in text-mining. Some use cases can
get by with article metadata (authors, title, etc.), some with
abstracts, but many use cases need full-text.

The landscape of access to full-text is a extremely hetergeous, with the
majority of variation along the publisher axis. The major hurdle are
paywalls. The majority of articles are published by the big three
publishers - Wiley, Springer, Elsever - and the majority of their
articles are behind paywalls.

A promising sign is that there's an increasing number of open access
publishers. xxxx.

\section{The discovery problem (maybe remove
section)}\label{the-discovery-problem-maybe-remove-section}

xxx

\section{Data sources}\label{data-sources}

There is increasing open source scientific literature content available
online. However, only a small proportion of scientific journals provide
access to their full content; whereas, most publishers provide open
access to their metadata only (most often through Crossref; Table 1).

Table 1. Sources of scientific literature, their content type provided
via web services, whether rOpenSci has an R packages for the service,
and where to find the API documentation.

\begin{longtable}[]{@{}llll@{}}
\toprule
Data Provider & Content Type & rOpenSci Pkg? & API
Documentation\tabularnewline
\midrule
\endhead
Crossref & Metadata only & rcrossref & \footnote{\url{http://api.crossref.org}}\tabularnewline
DataCite & Metadata only & rdatacite & \footnote{\url{https://support.datacite.org/docs/api}}\tabularnewline
Biodiversity Heritage Library & Full content/Metadata & rbhl &
\footnote{\url{http://bit.ly/KYQ1Rd}}\tabularnewline
Public Library of Science (PLoS) & Full text/altmetrics & rplos &
\footnote{\url{http://api.plos.org/solr}}\tabularnewline
Scopus (Elsevier) & Full content/Metadata & fulltext & \footnote{\url{http://bit.ly/J9S616}}\tabularnewline
arXiv & Full content/Metadata & aRxiv & \footnote{\url{https://arxiv.org/help/api/index}}\tabularnewline
Biomed Central (via Springer) & Full content/Metadata & fulltext &
\footnote{\url{https://dev.springer.com/}}\tabularnewline
bioRxiv & Full content/Metadata & fulltext & \footnote{\url{http://www.biorxiv.org/}}\tabularnewline
PMC/Pubmed (via Entrez) & Full content/Metadata & rentrez & \footnote{\url{https://www.ncbi.nlm.nih.gov/books/NBK25500}}\tabularnewline
Microsoft Academic Search & Metadata & fulltext/microdemic & \footnote{\url{https://azure.microsoft.com/en-us/services/cognitive-services}}\tabularnewline
\bottomrule
\end{longtable}

The following is a synopsis of the major data sources and associated R
tools.

\subsection{Crossref}\label{crossref}

Crossref is a non-profit that creates (or ``mints'') Digital Object
Identifiers (DOIs). In addition, they maintain metadata associated with
each DOI. The metadata ranges from simple (including author, title,
dates, DOI, type, publisher) to including number of citations to the
article, as well as references in the article, and even abstracts.

Crossref does have a text-mining opt-in program for publishers. The
result of this is that some publishers deposit URLs for full text
content of their articles. The majority of these links are pay-walled,
while some are open access. Using any of the various tools for working
with Crossref data, you can filter your search to get only articles with
full text links, and further to get only articles with full text links
that are open access.

The main interface for Crossref in R is
\href{https://github.com/ropensci/rcrossref}{rcrossref}. Parallel
interfaces are available in Ruby
(\href{https://github.com/sckott/serrano}{serrano}) and Python
(\href{https://github.com/sckott/habanero}{habanero}).

\subsection{Pubmed}\label{pubmed}

Pubmed is a corpus/website of NIH funded research \ldots{}
